---
title: "Tutorial"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message = FALSE}
library(EMBER)

library(dplyr)
library(ggplot2)

library(SummarizedExperiment)
library(msigdbr)
```

## Get PCA coordinates and plot

```{r}
# first calculate the normalization. by default it uses the 
# small PCA model
abim_100_normalized <- EMBER::get_final_ranking_values(
    sum_exp = abim_100, assay_to_use = "log2fpkm", 
    most_variable_genes = "small"
)

# calculate the embedding for the abim
abim_100_normalized <- EMBER::get_pca_coordinates(
    abim_100_normalized, which_pca = "small"
)


# get now the base plot so we can overlay the samples into the 
# embedding
base_plot <- EMBER::get_base_plot()

# plot on top of the base plot to have an idea where the samples
p <- base_plot + 
    ggplot2::geom_point(
        colData(abim_100_normalized) %>%
            data.frame,
        mapping = aes(x = PC3, y = PC4, color = pam50),
        size = 5,
        alpha = 1,
        shape = 17
    )
```

```{r}
p
```

All the samples are overlapping. We can see how close they 
are to the other cohorts in the space of batch effects.

```{r}
base_plot_batch <- EMBER::get_base_plot(
    x = "PC1", 
    y = "PC2",
    color = "cohort"
)

# plot on top of the base plot to have an idea where the samples
p_batch <- base_plot_batch + 
    ggplot2::geom_point(
        colData(abim_100_normalized) %>%
            data.frame,
        mapping = aes(x = PC1, y = PC2),
        color = "black",
        size = 3,
        alpha = 1,
        shape = 17
    )

p_batch
```

Makes sense since ABiM is a Swedish cohort that was obtained in a
similar way to SCANB.

## Regress data and calculate scores

Next step is to regress the data so we can calculate their scores.
We create a new summarized experiment that will contain the regressed
data, since only 8253 genes are regressed.

```{r}
# we first get the gene sets to calculate the scores
gene_sets <- EMBER::get_gene_sets(species = "Homo sapiens", category = "H")

abim_100_regressed <- EMBER::get_regressed_data(
    sum_exp = abim_100, assay_to_use = "log2fpkm"
) %>%
    # calculate the scores for the regressed data
    EMBER::get_scores_regressed(sum_exp = ., gene_sets = gene_sets)
```

If you check now the column names, there are 50 new columns with the
names "regressed_" and the name of the pathway. Moreover, one can also
check the regressed gene expression levels by checking the assay 
"regressed".

```{r}
hist(
        SummarizedExperiment::assay(abim_100_regressed, "regressed")["ESR1", ], 
        xlab = "Regressed ESR1 levels",
        main = "Histogram"
    )
```

## Comparing conditions

We can now compare the different conditions based on the pathway
scores obtained. More specifically, we compare ER status among the ABiM
100 samples.

```{r}
colData(abim_100_regressed) %>%
    data.frame %>%
    dplyr::filter(!is.na(ER)) %>%
    ggplot2::ggplot(aes(
        x = ER, 
        y = regressed_HALLMARK_ESTROGEN_RESPONSE_EARLY,
        color = ER
    )) + 
    ggplot2::geom_jitter() + 
    ggplot2::geom_boxplot(alpha = 0.5, outlier.shape = NA) + 
    ggplot2::labs(
        x = "ER status", 
        y = "Estrogen early",
        title = "Scores calculated on regressed data"
    ) + 
    ggplot2::theme_bw(base_size = 15)
```


